% "The Cost of a Steeper Link" — workshop-length paper.
% Facts source of truth: logs/FINDINGS_INVENTORY.md @ repo commit 1e7c13d.
%
% FORMAT NOTE: no NeurIPS .sty is vendored in this environment and the
% official style URL is not fetchable; per the paper-task spec this uses a
% clean generic article class with NeurIPS-like dimensions (10pt, Times,
% 5.5in x 9in text block). Swap the preamble block below for the venue
% .sty once the target venue is fixed.

\documentclass[10pt]{article}

% ---- NeurIPS-like layout (generic-substitute preamble) ----
\usepackage[verbose=true,letterpaper]{geometry}
\geometry{textheight=9in,textwidth=5.5in,heightrounded}
\usepackage{newtxtext,newtxmath}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{booktabs,graphicx,amsmath} % amssymb omitted: newtxmath provides those symbols and clashes on \Bbbk
\usepackage[numbers,sort&compress]{natbib}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\bibliographystyle{plainnat}

\newcommand{\mpd}{\ensuremath{\times}MPD}

\title{The Cost of a Steeper Link: Calibration, Stability, and Tie
Mechanics in Large-Scale Pairwise Preference Ranking}

\author{%
  Aviral Poddar\thanks{Author list and affiliations placeholder — to be
  finalized.}\\
  \texttt{aviral@example.org}
}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Large-scale preference leaderboards such as Chatbot Arena rank language
models by maximum-likelihood Bradley--Terry (BT), inheriting the logistic
link's independence-of-irrelevant-alternatives assumption over a comparison
pool that never stops growing. We test whether a flexible Thurstonian
alternative---a lattice-based model that on pairwise data reduces to a
steeper, natively tie-aware gap link---measurably improves refit
stability, held-out calibration, or tie handling on 1.67 million Arena
votes spanning 129 models and sixteen months. Every comparison was
pre-registered: effect-size thresholds derived before fitting, synthetic
detection gates passed before real data, verdict criteria fixed in
advance. We find no demonstrable advantage in either direction. Refit
stability under population growth is statistically indistinguishable
(median paired rank-correlation difference $+0.00000$ across thirteen
month-over-month refits); two structurally different tests agree that link
shape is not discriminable at the ability gaps real matchups occupy
(analytic ceilings of $0.23$--$0.31\times$ our minimum practical
difference); and the calibration and tie-mechanism comparisons both end
\emph{inconclusive}, each dominated by a single-window episode rather than
any systematic difference. Inside this null sit two structural findings: a
steeper link's single tie parameter entangles tie mass with decisive
discrimination, whereas Davidson's tie parameter is provably orthogonal to
its logistic link; and under parameter uncertainty, steep links trade
prediction-side overconfidence against an estimation-side shrinkage that
once protected against a severe cold-start extrapolation error. One
anomaly survives every attribution attempt: temporally localized excess
dispersion under every link tested. Practically, cold-start coverage and
nonstationary drift---not link choice---bind next-month predictability in
deployments of this shape.
\end{abstract}

\section{Introduction}\label{sec:intro}

Chatbot Arena~\citep{chiang2024arena} has become the de facto public
scoreboard for large language models: users pose a prompt, two anonymous
models respond, and the user votes for a winner or declares a tie. The
platform converts these votes into a leaderboard by maximum-likelihood
Bradley--Terry (BT)~\citep{bradleyterry1952}, fitting one latent ability
per model under a logistic link. Our study window alone contains 1.8
million recorded battles across 129 models (1.67 million after the
platform's own deduplication filter), with new models entering the
comparison pool continuously throughout.

BT buys its simplicity with a strong structural assumption. As a
logit-family choice model it satisfies Luce's independence of irrelevant
alternatives~\citep{luce1959}: the odds of model $A$ beating model $B$ are
taken to be invariant to whatever else occupies the pool. Nothing about
human preference data guarantees this, and a leaderboard whose pool grows
by the month---forcing a global re-estimate over an ever-changing
comparison graph---is precisely the setting where one might expect the
assumption to bite: incumbent ratings could shift when entrants arrive,
predicted win probabilities could miscalibrate for fresh models, and the
tie outcome (roughly a fifth of quality votes) is not even part of BT's
native outcome space. The Thurstonian random-utility
family~\citep{thurstone1927} offers the classical alternative---each model
draws a latent performance and the better draw wins---and Cotton's
lattice construction~\citep{cotton2021} makes that family computationally
exact for arbitrary base densities, with ties arising natively as
dead-heat mass rather than by parametric extension~\citep{davidson1970}.
On pairwise data the lattice model reduces to a \emph{gap link}: a
probability curve in the ability difference, like BT's logistic but with
a different shape---steeper at a toss-up (slope $0.284$--$0.337$
depending on its tie-band width, against the logistic's $0.25$) and
carrying a native tie band. Whether that different shape is worth
anything in a real deployment is an empirical question, and to our
knowledge it had not been asked at this scale.

We asked it four ways, each pre-registered before any real-data number
was computed: practical-significance thresholds derived from
deployment-meaningful error sizes, detection gates validated on synthetic
ground truth, and pattern-based verdict criteria fixed in advance.
(RQ1)~When new models enter, whose refit perturbs incumbent rankings
less? (RQ2)~Is either link shape more consistent with observed win rates
across triples of models? (RQ3)~Which link calibrates better on held-out
votes, especially for recent entrants? (RQ4)~Does a native dead-heat band
fit the observed tie structure better than Davidson's extension of BT?

The answer, delivered with unusual statistical redundancy, is that the
link does not matter here. Refit stability is indistinguishable---across
thirteen month-over-month refits the median paired rank-correlation
difference is $+0.00000$, with the two methods exactly tied in five
windows---while window-to-window variation in stability, governed by pool
maturity, exceeds the between-method differences by roughly two orders of
magnitude. Two structurally different analyses (an
analytic calibration ceiling and an empirical triple-additivity test)
independently agree that at the ability gaps real matchups occupy, no
plausible member of the Thurstonian family is distinguishable from the
logistic. The two remaining comparisons, held-out calibration and tie
handling, both end \emph{inconclusive} rather than equivalent---a
distinction we maintain deliberately, because each verdict was dominated
by a single anomalous window rather than by any systematic difference,
and our pre-registered criteria refuse a call under exactly that pattern.

This paper's contribution is that negative result, stated with the
precision it deserves, together with what we found inside it. First, a
structural asymmetry: a steep link's single tie parameter entangles tie
mass with decisive discrimination, while Davidson's tie parameter is
provably orthogonal to its logistic link---forcing the lattice band to
absorb both-bad-scale tie mass measurably degrades its decisive
accuracy. Second, a two-faced interaction between link steepness and
parameter uncertainty: steep links make overconfident plugin predictions
from noisy estimates, yet the same steepness acts as implicit shrinkage
in estimation---in our data it once prevented a severe cold-start
extrapolation error that cost BT dearly. Third, an open question: a
temporally localized excess dispersion, present under every link we
tested and attributable to none of the covariates we measured. And
throughout, a deployment lesson that inverts the motivating question:
what binds ranking quality in a system like Arena is not the link
function at all, but cold-start coverage and nonstationary drift---both
of which dwarf every between-method difference we measured. We believe
the audit itself, and the pre-registration discipline that kept four
successive verdicts honest, are as much the contribution as any single
number.

% ==================================================================
% Sections to be added one at a time, each user-reviewed before the next:
%   3. Related Work
%   4. Data and Methods
%   5. RQ1  Refit Stability            (Fig 1: rq1_metrics.csv)
%   6. RQ2b Link Shape / Additivity    (Fig 2: rq2b_excess_covariates.csv)
%   7. RQ3  Held-Out Calibration
%   8. RQ4  Tie Mechanisms             (Fig 3: rq4_tie_curve_bins.csv + bands)
%   9+ (spec pending from user)
% ==================================================================

\bibliography{refs}

\end{document}
